---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Install the needed packages and after ensuring the data is in our working directory, we read the data and verify the dimensions. 
```{r echo = TRUE, message = FALSE}
library(ggplot2)
library(dplyr)
data1 <- read.csv(unzip("activity.zip"))
dim(data1)

```


## What is mean total number of steps taken per day?

```{r echo = TRUE, message = FALSE, warning = FALSE}
# summarize the total steps for each date
data2= group_by(data1, date)
# if total=sum(steps, na.rm = TRUE), total will be 0 instead of NA
data2 = summarize(data2, total=sum(steps))

# make a histogram
ggplot(data=data2) + geom_histogram(mapping = aes(total))+
  labs(title="Histogram of total steps each day",  x = "Total Steps")

# calculate the mean and median total daily steps
mean(data2$total, na.rm=TRUE)
median(data2$total, na.rm=TRUE)
```
The mean is 10766.19 and the median is 10765.

## What is the average daily activity pattern?

```{r echo = TRUE, message = FALSE, warning=FALSE}
# calculate the mean for each interval
data3= group_by(data1, interval)
data3 = summarize(data3, meansteps=mean(steps, na.rm=TRUE))

# make a time series plot
ggplot(data3) + geom_line(aes(interval,meansteps), type = "l") +
  labs(title = "Average Daily Activity Pattern", x = "Interval Representing Time of Day", y = "Average Steps")
```


```{r echo = TRUE, message = FALSE, warning = FALSE}
# Find the interval with the maximum meansteps value in data3 (grouped by intervals)
data3[which.max(data3$meansteps),]

```
Interval 835 has the highest meansteps value.

## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

```{r echo = TRUE, message = FALSE, warning = FALSE}
 missing.steps <- sum(is.na(data1$steps))
    paste("The number of missing values in the dataset is",missing.steps)
```
The number of missing values in the dataset is 2304.

We replace any NAs in steps with the mean for the corresponding interval.
```{r echo = TRUE, message = FALSE, warning = FALSE}
# add the mean interval values as a new column to the original data
data4 = left_join(x = data1, y = data3, by = "interval")
dim(data4)
# replace the mean interval value if steps==NA
data4$steps[is.na(data4$steps)] <- data4$meansteps[is.na(data4$steps)]
```

Now the steps column in data4 has the missing values filled in.

```{r echo = TRUE, message = FALSE, warning = FALSE}
# summarize the total steps for each date
data5= group_by(data4, date)
# if total=sum(steps, na.rm = TRUE), total will be 0 instead of NA
data5 = summarize(data5, total=sum(steps))

# make a histogram
ggplot(data=data5) + geom_histogram(mapping = aes(total))+
  labs(title="Histogram of total steps each day with imputed missing values",  x = "Total Steps")

# calculate the mean and median total daily steps
mean(data5$total, na.rm=TRUE)
median(data5$total, na.rm=TRUE)
```
The mean is 10766.19 and the median is now 10766.19.The mean remained the same, but there is a modest change in the median value.

## Are there differences in activity patterns between weekdays and weekends?

```{r echo = TRUE, message = FALSE, warning = FALSE}
# Create a day column and then a weekend column
data4$day = weekdays(as.Date(data4$date))
data4$weekend = ifelse(data4$day=="Saturday"|data4$day=="Sunday","Weekend","Weekday")

# calculate the mean for each interval, for both weekends and weekdays
data6= group_by(data4, weekend, interval)
data6 = summarize(data6, meansteps=mean(steps, na.rm=TRUE))

# make a time series plot
ggplot(data6) + geom_line(aes(interval,meansteps), type = "l") +
  facet_grid(weekend ~ .) +
  labs(title = "Average Daily Activity Pattern", x = "Interval Representing Time of Day", y = "Average Steps")

```

